Dependent Object Types

### 2012-09-11 doc/candidacy/slides.tex
system Fsub
ml section rev

\begin{frame}{SML modules}
\begin{description}[type-sharing]
\item[structure] a module\\a collection of types, values and
  (sub)structures packaged as a program unit.
\item[signature] an interface; the type of a structure\\description of
  the types, values and (sub)structures of a structure given by their
  kinds, types and interfaces.
\item[functor] a function mapping structures to structures
\item[type-sharing] enables stating that types specified in separate
  modules are actually equal
\end{description}
\end{frame}

\begin{frame}[fragile]{SML modules: Example}
\inputminted[fontsize=\footnotesize]{sml}{intro.sml}
\end{frame}

@@

\begin{frame}{ML-Like: Calculus}
grounded in type theory; based on Girard's $F_\omega$, adding:
\begin{itemize}
\item {\it translucent sums} to model modules
\item dependent functions to model functors
\item a notion of subtyping to model module implementation-interface matching
\end{itemize}
\end{frame}

\begin{frame}{ML-Like: Translucent Sums}
\begin{itemize}
\item translucent sums are values representing modules
\item translucent sum: sequence of bindings
\item translucent sum type: sequence of declarations
\item unlike traditional records, later fields can depend on earlier ones
\item fields can be types or terms
\item any type can be partially or fully determined by a type expression
\item dependent typing since translucent sums are terms that may contain type components
\end{itemize}
\end{frame}

\begin{frame}[fragile]{ML-Like: Example}
\begin{minted}{sml}
structure XInt = struct
  type T = int
  val v = 3
  val f = negate
end
structure XBool = struct
  type T = bool
  val v = true
  val f = not
end
structure X =
  if flip() then XInt else XBool
signature XType = sig
  type T
  val v: T
  val f: T -> T
end
\end{minted}
\end{frame}

% Maintaing Soundness?

\begin{frame}{ML-Like: Path-Dependent Types}
\begin{itemize}
\item Scala's abstract types  $\approx$ ML's abstract types of signatures; limitations:
\begin{itemize}
\item recursive references
\item bounded quantifications
\end{itemize}
\item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for %example with justification for two items below? not clear without introducing internal names?
\begin{itemize}
\item breaking the dependencies between (sub)fields
\item propagating typing information
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{System $F_\omega$: Higher-order polymorphic lambda-calculus}
\begin{description}[1234567890]
\item[terms $t$]
\begin{description}[123456789012345678901234567890]
\item[typed lambda-calculus terms] $x$ -- $\lambda x:T.t$ -- $t t$
\item[type abstraction] $\lambda X::K.t$
\item[type application] $t [T]$
\end{description}
\item[types $T$]
\begin{description}[123456789012345678901234567890]
\item[type variable] $X$
\item[type of functions] $T \rightarrow T$
\item[universal type] $\forall X::K.T$
\item[operator abstraction] $\lambda X::K.T$
\item[operator application] $T T$
\end{description}
\item[kinds $K$]
\begin{description}[123456789012345678901234567890]
\item[kind of proper types] $*$
\item[kind of operators] $K \Rightarrow K$
\end{description}
\end{description}
\end{frame}

% System F sub
\end{document}

>>>>

\begin{frame}[fragile]{SML modules: Structures and Signatures}
\begin{columns}
\begin{column}[t]{5cm}
\begin{minted}{sml}
structure A = struct
  type T = int
  val v = 0
  val f = fn x => x + 1
end

signature X = sig
  type T
  val v: T
  val f: T -> T
end

structure B :> X = A
\end{minted}
\end{column}
\begin{column}[t]{5cm}
\begin{minted}{sml}
(* A.T transparent *)
val vA: int = A.v

(* B.T opaque *)
val vB: B.T = B.v
\end{minted}
\end{column}
\end{columns}
\end{frame}

\begin{frame}[fragile]{SML modules: Functions and Type Sharing}
\begin{minted}{sml}
functor mkDoubler(arg: X) :> X where type T = arg.T = struct
  type T = arg.T
  val v = arg.v
  val f = arg.f o arg.f
end

(* A2.T = A.T = int *)
structure A2 = mkDoubler(A);
val a: int = A2.f(A.v);

(* B2.T = B.T *)
structure B2 = mkDoubler(B);
val b: B.T = B2.f(B.v);
\end{minted}
\end{frame}

\begin{frame}{SML modules: Recap}
\begin{description}[type-sharing]
\item[structure] a module\\a collection of types, values and
  (sub)structures packaged as a program unit.
\item[signature] an interface; the type of a structure\\description of
  the types, values and (sub)structures of a structure given by their
  kinds, types and interfaces.
\item[functor] a function mapping structures to structures
\item[type-sharing] enables stating that types specified in separate
  modules are actually equal
\end{description}
\end{frame}

@@

\begin{frame}{``Translucent'' ML: Calculus}
grounded in type theory; based on Girard's $F_\omega$, adding:
\begin{itemize}
\item {\it translucent sums} to model modules
\item dependent functions to model functors
\item a notion of subtyping to model module implementation-interface matching
\end{itemize}
\end{frame}

\begin{frame}{``Translucent'' ML: Translucent Sums}
\begin{itemize}
\item translucent sums are values representing modules
\item translucent sum: sequence of bindings
\item translucent sum type: sequence of declarations
\item unlike traditional records, later fields can depend on earlier ones
\item fields can be types or terms
\item any type can be partially or fully determined by a type expression
\item dependent typing since translucent sums are terms that may contain type components
\end{itemize}
\end{frame}

\begin{frame}[fragile]{``Translucent'' ML: Example}
\begin{minted}{sml}
structure XInt = struct
  type T = int
  val v = 3
  val f = negate
end
structure XBool = struct
  type T = bool
  val v = true
  val f = not
end
structure X =
  if flip() then XInt else XBool
signature XType = sig
  type T
  val v: T
  val f: T -> T
end
\end{minted}
\end{frame}

\begin{frame}{``Translucent'' ML: Path-Dependent Types}
\begin{itemize}
\item Scala's abstract types  $\approx$ ML's abstract types of signatures; limitations:
\begin{itemize}
\item recursive references
\item bounded quantifications
\end{itemize}
\item translucent sum can be given a more precise type by referring to the name of its type member with a path selection; essential for %example with justification for two items below? not clear without introducing internal names?
\begin{itemize}
\item breaking the dependencies between (sub)fields
\item propagating typing information
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]{System $F_\omega$: Higher-order polymorphic lambda-calculus}
\begin{description}[1234567890]
\item[terms $t$]
\begin{description}[123456789012345678901234567890]
\item[typed lambda-calculus terms] $x$ -- $\lambda x:T.t$ -- $t\ t$
\item[type abstraction] $\lambda X::K.t$
\item[type application] $t [T]$
\end{description}
\item[types $T$]
\begin{description}[123456789012345678901234567890]
\item[type variable] $X$
\item[type of functions] $T \rightarrow T$
\item[universal type] $\forall X::K.T$
\item[operator abstraction] $\lambda X::K.T$
\item[operator application] $T\ T$
\end{description}
\item[kinds $K$]
\begin{description}[123456789012345678901234567890]
\item[kind of proper types] $*$
\item[kind of operators] $K \Rightarrow K$
\end{description}
\end{description}
\end{frame}

\begin{frame}[fragile]{System $F_{<:}$: Bounded Quantification}
\begin{description}[1234567890]
\item[terms $t$]
\begin{description}[123456789012345678901234567890]
\item[typed lambda-calculus terms] $x$ -- ${\lambda}x:T.t$ -- $t\ t$
\item[type abstraction] $\lambda X<:T.t$
\item[type application] $t [T]$
\end{description}
\item[types $T$]
\begin{description}[123456789012345678901234567890]
\item[type variable] $X$
\item[maximum type] $\top$
\item[type of functions] $T \rightarrow T$
\item[universal type] $\forall X<:T.T$
\end{description}
\end{description}
\end{frame}

\end{document}
